---
title: "Homework 7"
author: "Mckenzie Scriven"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run.  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Data 
For this assignment, we will use data from a study on elephants and the effects of poaching on tusk size.  

Reference: [Chiyo, Patrick I., Vincent Obanda, and David K. Korir. “Illegal tusk harvest and the decline of tusk size in the African elephant.” Ecology and Evolution 5, 22: 5216–5229 (2015)](https://doi.org/10.1002/ece3.1769). Data deposited at [Dryad Digital Repository](https://doi.org/10.5061/dryad.h6t7j).  

**1. Before starting data analysis, read the abstract of the paper to get an idea of the questions being asked. In 2-3 sentences, describe what the study is testing and the variables involved.**  
Study is testing whether hunting affects phenotype of tusks in elephants.
Variables include tusk length, tusk circumference, soulder height, age, and date sampled

**2. Load `elephants.csv` and store it as a new object called `elephants`.**  
```{r}
elephants <- read.csv("data/elephants.csv")
```

**3. Clean the data by converting variable names to lowercase with no spaces or special characters.**  
```{r}
elephants <- clean_names(elephants)
```


**4. Use one or more of the summary functions you have learned to get an idea of the structure of the data.**   
```{r}
glimpse(elephants)
```


**5. Use `mutate()` Change the variables `years_of_sample_collection`, `elephant_id`, and `sex` to factors. Be sure to store the output as a new dataframe and use it for the remaining questions.**  
```{r}
elephants <- elephants %>% 
  mutate(years_of_sample_collection=as.factor(years_of_sample_collection),
         elephant_id=as.factor(elephant_id),
         sex=as.factor(sex))
```


**6. From which years were data collected? Show the sample periods below.**  
```{r}
elephants %>% 
  select(years_of_sample_collection) %>% 
  distinct() 
```


**7. How many males and females were sampled in this study?**  
```{r}
elephants %>% 
  count(sex)
```

**8. What is the mean, median, and standard deviation for age of males and females included in the study? Separate the results by year of sample collection. Does the sampling look even between years and sexes?**  
males 1966-68:
```{r}
elephants %>% 
  filter(years_of_sample_collection=="1966-68", sex=="m") %>% 
  summarize(mean=mean(estimated_age_years), med=median(estimated_age_years), sd=sd(estimated_age_years))
```
females 1966-68:
```{r}
elephants %>% 
  filter(years_of_sample_collection=="1966-68", sex=="f") %>% 
  summarize(mean=mean(estimated_age_years, na.rm=T), med=median(estimated_age_years, na.rm=T), sd=sd(estimated_age_years, na.rm=T))

```
females 2005-13:
```{r}
elephants %>% 
  filter(years_of_sample_collection=="2005-13", sex=="f") %>% 
  summarize(mean=mean(estimated_age_years), med=median(estimated_age_years), sd=sd(estimated_age_years), )

```
Males from 2005-13:
```{r}
elephants %>% 
  filter(years_of_sample_collection=="2005-13", sex=="m") %>% 
  summarize(mean_age_m=mean(estimated_age_years, na.rm=T), med=median(estimated_age_years, na.rm=T), sd=sd(estimated_age_years, na.rm=T))

```
in 1966-68, the males are much younger than the females. in 2005-13, the females are still older than the males on average, but by much less when looking at the mean. The median for the males in 2005-13 is much lower than the mean, which suggests that there are a few very old males who are skewing the data and most males are still young.

**9. Is age (independent variable) a positive predictor of tusk length (dependent variable)? Create a plot that shows the relationship between these variables and add a linear model fit line.**  
```{r}
elephants %>% 
  ggplot(mapping=aes(x=estimated_age_years,y=tusk_length_in_cm))+
  geom_point()+
  geom_smooth(method=lm)+
  labs(title="Age vs tusk length", x="Estimated age (years)",y="Tusk length (cm)")
```


**10. Is shoulder height (independent variable) a positive predictor of tusk length (dependent variable)? Create a plot that shows the relationship between these variables and add a linear model fit line.**  
```{r}
elephants %>% 
  ggplot(mapping=aes(x=shoulder_height_in_cm,y=tusk_length_in_cm))+
  geom_point()+
  geom_smooth(method=lm)+
  labs(title="Shoulder height vs tusk length", x="Shoulder height (cm)",y="Tusk length (cm)")

```


**11. The authors argue that because poachers preferentially target elephants with large tusks, this has resulted in a decrease in average tusk length. Is this supported by the data? Show your code and calculations below.**  
```{r}
elephants %>% 
  filter(years_of_sample_collection=="2005-13") %>% 
  summarize(mean_length_twentythirteen=mean(tusk_length_in_cm, na.rm=T), med=median(tusk_length_in_cm, na.rm=T), sd=sd(tusk_length_in_cm, na.rm=T))

elephants %>% 
  filter(years_of_sample_collection=="1966-68") %>% 
  summarize(mean_length_sixties=mean(tusk_length_in_cm, na.rm=T), med=median(tusk_length_in_cm, na.rm=T), sd=sd(tusk_length_in_cm, na.rm=T))

```
The mean and median length for all elephants in 2005-13 is much lower than the mean and median length for all elephants in 1966-68, so their claim is supported. 

**12. Male elephants reach effective sexual maturity at 25 years while females are sexually mature at 12 years. Make a new dataframe that extracts only the males and females at sexual maturity. Then, make a plot that shows the range of tusk length between the two sample periods for these mature elephants.**   
```{r}
elephants %>% 
  filter((sex=="m" & estimated_age_years>=25) | (sex=="f" & estimated_age_years>=12)) %>% 
  ggplot(mapping=aes(x=years_of_sample_collection,y=tusk_length_in_cm))+
  geom_boxplot(fill="pink")+
  labs(title="Tusk length in 1966-68 and 2005-13 in mature elephants",x="Years sampled",y="Tusk length (cm)")
```

## Submit the Homework
1. Save your work and knit the .rmd file.  
2. Open the .html file and "print" it to a .pdf file in Google Chrome (not Safari).    
3. Go to the class Canvas page and open Gradescope.  
4. Submit your .pdf file to the homework assignment- be sure to assign the pages to the correct questions.  
5. Commit and push your work to your repository. 